Rotations and Insecticide Resistance the story

### 2018-01-26 inst/documents/rotations_notes.Rmd
improve vis of insecticide use

From GPIRM :
Idea that cost of resistance likely to decline when ferquency of resistance gene has been high. (We don't have this in our models).
Insecticide resistance management strategies must, however, be implemented before the resistance gene becomes common and stable in the population; otherwise, the resistant gene will not recede even if use of the insecticide causing selection pressure is discontinued. As demonstrated by a study of blowflies by McKenzie and Whitten in 1982 (3), fitness cost is not an intrinsic property of the gene. Therefore, if that gene is allowed sufficient time to become common in a population, the rest of the genome will adapt to incorporate it without a significant fitness cost. At this point, even if the selection pressure is removed, the resistance gene will remain in the population.

@@

@@

@@

>>>>

From GPIRM :
Idea that cost of resistance likely to decline when frequency of resistance gene has been high. (We don't have this in our models).
Insecticide resistance management strategies must, however, be implemented before the resistance gene becomes common and stable in the population; otherwise, the resistant gene will not recede even if use of the insecticide causing selection pressure is discontinued. As demonstrated by a study of blowflies by McKenzie and Whitten in 1982 (3), fitness cost is not an intrinsic property of the gene. Therefore, if that gene is allowed sufficient time to become common in a population, the rest of the genome will adapt to incorporate it without a significant fitness cost. At this point, even if the selection pressure is removed, the resistance gene will remain in the population.

But I looked at the 1982 paper and not sure it supports that ?
Seems to me that McKenzie1982 shows that relative fitness of the susceptible genotypes increase over time as would be expected by a decline in the concentration of the active ingredient ? (and similar to our WoS paper)

@@

*22/1/2018*

Why do rotations last longer than sequences ? (even when coverage=1 & cost=0)

I think it's just because the rotation interval continues to its end.

e.g. added an option to plot a 2nd scenario to rot_plot_resistance()

dfrot0 <- run_rot(n_insecticides=2, rot_interval=0, expo_hi=0.5, eff=0.5, cost=0, coverage=1)
dfrot15 <- run_rot(n_insecticides=2, rot_interval=15, expo_hi=0.5, eff=0.5, cost=0, coverage=1)
rot_plot_resistance(dfrot0, df_resanother=dfrot15, plot_refuge=FALSE)

So maybe we should have assessment of the resistance frequency in the rotation strategy too ?
I could add it as a flag.

done, now works for simple scenario

BUT rotations still lasting a few generations longer in later scenarios ?
dfseq <- run_rot(n_insecticides=2, rot_interval=0, expo_hi=0.5, eff=0.5, cost=0.01, coverage=1)
dfrot <- run_rot(n_insecticides=2, rot_interval=15, expo_hi=0.5, eff=0.5, cost=0.01, coverage=1)
rot_plot_resistance(dfrot, df_resanother=dfseq, plot_refuge=FALSE)

rotation
insecticide2 192 2 active 0.46672149
insecticide2 193 2 active 0.48524932
insecticide2 194 NA active 0.50379224

sequence
insecticide1 187 0.4710576
insecticide1 188 0.4895936
insecticide1 189 0.5081326

Could it be because an extra generation is slipped in at each rotation switchover ?
Or might be that more decline happens in rotations

Actually I think it's due to min_rwr_interval which was set at 5 by default and in the current code is implemented for sequence but not for rotation.

hmmm but even when I set min_rwr_interval to 1 there seems to be the difference ?
is there any info in the console messages ?

###rotation :
generation 166, switch from insecticide 1 to 2; frequencies = 0.402193 and 0.178446
generation 181, switch from insecticide 2 to 1; frequencies = 0.402193 and 0.384221
generation 188, switch from insecticide 1 to 2; frequencies = 0.512362 and 0.393770

simulation terminating at generation 194 because all RAFs above threshold of 0.500000

frequency of resistance in females to insecticide 1 is 0.504826
frequency of resistance in females to insecticide 2 is 0.503792

###sequence :
generation 92, switch from insecticide 1 to 2; frequencies = 0.510798 and 0.001000
generation 183, switch from insecticide 2 to 1; frequencies = 0.510798 and 0.397957

simulation terminating at generation 189 because all RAFs above threshold of 0.500000

frequency of resistance in females to insecticide 1 is 0.508133
frequency of resistance in females to insecticide 2 is 0.503261

Hi Ian,

I'm looking into rotations and mostly I find very little difference from sequences even with costs or refuges. In a sequence resistance builds up once and then declines, in rotations it builds up in steps and declines in between but the time-to-resistance is the same for both.

Then I refound this idea in GPIRM that costs of resistance are likely to decline when frequency of resistance gene has been high.

"Insecticide resistance management strategies must, however, be implemented before the resistance gene becomes common and stable in the population; otherwise, the resistant gene will not recede even if use of the insecticide causing selection pressure is discontinued. As demonstrated by a study of blowflies by McKenzie and Whitten in 1982 (3), fitness cost is not an intrinsic property of the gene. Therefore, if that gene is allowed sufficient time to become common in a population, the rest of the genome will adapt to incorporate it without a significant fitness cost. At this point, even if the selection pressure is removed, the resistance gene will remain in the population."

If this was the case it would probably favour rotations over sequences.

Do you have any thoughts ? Do you think it likely that costs will decline over time at high resistance frequencies ?

I looked at the 1982 paper (attached) and I don't think it supports the idea that costs decline over time (but it does seem to have data on relative fitness of RR,RS,SS that we could use in the WoS paper, gives evidence for dominance changing over time and could be used to derive our model parameters.)

Andy

Thanks Andy

That results should put the cat amongst the pigeons i.e. little difference between sequences and rotations.

There are examples of “compensatory” mutations offsetting the fitness cost of a mutation. Often from bacteria where populations can be very high and the population is asexual so that the mutation and modifier are always linked. It is less well studied in larger animals. There may be examples from IR but I am unaware. Often it is a post hoc explanation plucked out the air to explain why there is no apparent fitness effect i.e. “there were fitness effect, but they must have been modified”.

I like the 1982 paper… I have put it in the WoS folder with a note to discuss it.

*25/1/2018*

noticed that rotation plots don't have a log y axis where resistance plots do.
There is a logy option in run_rot() should probably set that TRUE by default.
yes fixed

Rowland(1981) has some good stuff on costs of resistance and selection coefficients and putting these into a model and testing against field data :

Withdrawal of dieldrin in two regions of India led to the frequency of SS increasing from 0.31 to 0.62 within 7 months (Rao et al., 1960) and from 0.11 to 0.85 within 9 months (Bhatia & Deobhankar, 1963).
In Fig. 9 the computed reversions in An.garnbiae and Anstephemi have been redrawn, expressed in terms of SS rather than R frequency. Superimposed on these are the An. culicifacies field reversions. Assuming that one generation takes 1 month (cf. Curtis et al., 1978), the computer
predictions seem to be consistent with the field reversions.

Having established why and to what extent resistance should revert when cyclodiene usage is ended, the question remains could yHCH be effectively re-introduced at a later date as part of a insecticide-rotation resistance management strategy? Because RR is the most deleterious genotype, the rate of reversion should be rapid initially. Only when R is predominantly carried by heterozygotes would the rate of reversion begin to slow down. For this reason resistance cannot be expected to revert completely, but it should nevertheless revert sufficiently to warrant the useful re-introduction of yHCH for several generations at a time. An unresolved question, potentially damagingto the long-term future of the rotation strategy,is would modifiers of fitness evolve if cyclodieneusage was prolonged?

It seems, then, that rotation strategies might prolong dieldrin efficacy indefinitely, not just against anophelines but against many species of insect.

[because the combination of selection coefficient, cost and dominance means that resistance doesn't rise too fast when used and declines when not in use].

Rowland2013 data on hut trials of actellic - pirimphos-methyl - an OP. Includes monthly change and at diff concentrations so could be useful for WoS, but doesn't have resistant & susceptible (does have resistance frequencies).

Matowo2015 selection of insecticide resistance, seems to show changing insecticide resistance in response to interventions. but only shows mortality change not resistance frequency change.

Kim2016 interesting risk paper :
The main policy implication of this analysis is that evolutionary speed of insecticide resistance and
maximal larviciding efficacy strongly affect the preferred actions implied by our decision analysis tool and that these parameters jointly offer the largest potential benefits from reduced uncertainty

*26/1/2018*

what to call new UI resistrot consistent with resistmixseq

@@

Explore idea from Rowland (1981) that resistance declines faster when it is mostly held by the RR (i.e. at higher allele frequencies)

Look into data on changing allele frequencies over time (e.g. for Kdr)...

It occurred to me that maybe a benefit of rotations is that if it is difficult to measure resistance allele frequencies they provide a lower risk of reaching resistance thresholds for any insecticide. i.e. are they a better strategy in light of uncertainty in resistance allele frequencies ?

~ check again on current small differences between rotations & sequences

~ maybe I can write a function to plot 2 scenarios on the same graph (in the same way that curtis fig2 plots mix & seq on same graph) ?
~ I could have the insecticide use still along the top but with diff colours showing the two alternatives

@@

~ can we get cost to decrease over time at high frequency as suggested in GPIRM ?

### 2018-01-26 inst/documents/rotations_story.Rmd
improve vis of insecticide use

If there are no costs of resistance or migration from untreated areas then rotations and sequences are likely to result in a similar effect on resistance frequencies as seen in the following plots.

*Question to myself* why do rotations last c 5 generations longer in all comparisons ?

@@

dfr <- run_rot(n_insecticides=2, rot_interval=0, expo_hi=0.5, eff=0.5, cost=0, coverage=1)

@@

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="A rotation, insecticide changed every 15 generations (c 1 year)."}

dfr <- run_rot(n_insecticides=2, rot_interval=15, expo_hi=0.5, eff=0.5, cost=0, coverage=1)

@@

```{r, eval=FALSE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="A rotation, insecticide changed every 45 generations (c 3 years)."}
#eval=FALSE
dfr <- run_rot(n_insecticides=2, rot_interval=45, expo_hi=0.5, eff=0.5, cost=0, coverage=1)

@@

\newpage
If there are costs of resistance then we would expect resistance frequencies to decline when an insecticide is not in use. However this may not make much difference between the effect of a sequence and a rotation. In a sequence the resistance frequency builds up once and then gets a long period to decline when not in use. In a rotation resistance frequency increases as a series of small steps with short declines in between (see following plots).

(This is assuming that costs don't change according to the frequency of the resistance allele - there is some evidence to suggest that costs may decline when a resistance allele has been at a high frequency for a long time).

@@

dfr <- run_rot(n_insecticides=2, rot_interval=0, expo_hi=0.5, eff=0.5, cost=0.01, coverage=1)

@@

dfr <- run_rot(n_insecticides=2, rot_interval=15, expo_hi=0.5, eff=0.5, cost=0.01, coverage=1)

@@

\newpage
Even without costs of resistance we might expect resistance frequencies to decline when an insecticide is not in use due to immigration from nearby areas that were not treated originally. This is shown in the following two plots where a neighbouring untreated area and exchange of migrants is represented.

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="A sequence with dispersal between treated and untreated areas (no resistance costs)."}

dfr <- run_rot(n_insecticides=2, rot_interval=0, expo_hi=0.5, eff=0.5, cost=0, coverage=0.8, migration = 0.01)

@@

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="A 1 year rotation with with dispersal between treated and untreated areas (no resistance costs)."}

dfr <- run_rot(n_insecticides=2, rot_interval=15, expo_hi=0.5, eff=0.5, cost=0, coverage=0.8, migration = 0.01)

@@

>>>>

If there are no costs of resistance or migration from untreated areas then rotations and sequences are likely to result in very similar effect on resistance frequencies as seen in the following plots.

These plots assume that an insecticide is changed when it reaches the resistance frequency threshold even if the rotation interval has yet to be reached. Without this assumption then rotations always take a little longer time than sequences but at the expense of reaching slightly higher resistance frequencies.

The plots here are for 2 insecticides to keep things simple however the model can be run for any number of insecticides. Effectiveness of insecticides at killing SS is set to 0.8 unless indicated.

@@

dfrot0 <- run_rot(n_insecticides=2, rot_interval=0, expo_hi=0.5, eff=0.8, cost=0, coverage=1)

@@

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="A rotation, insecticide changed every 10 generations (c 1 year)."}

dfrot15 <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=10, expo_hi=0.5, eff=0.8, cost=0, coverage=1)

@@

```{r, eval=FALSE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="A rotation, insecticide changed every 30 generations (c 3 years)."}
#eval=FALSE
dfr <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=30, expo_hi=0.5, eff=0.8, cost=0, coverage=1)

@@

\newpage
###costs of resistance

If there are costs of resistance then we would expect resistance frequencies to decline when an insecticide is not in use. However this may not make much difference between the effect of a sequence and a rotation. In a sequence the resistance frequency builds up once and then gets a long period to decline when not in use. In a rotation resistance frequency increases as a series of small steps with short declines in between (see following plots). The time it takes for both insecticides to reach the resistance threshold of 50% is still very similar for a sequence and a rotation.

(This is assuming that costs don't change according to the frequency of the resistance allele - there is some evidence to suggest that costs may decline when a resistance allele has been at a high frequency for a long time (mentioned in GPIRM), if this is the case then it could indeed provide an advantage for rotations over sequences).

@@

dfseq <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=0, expo_hi=0.5, eff=0.8, cost=0.01, coverage=1)

@@

dfrot <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=10, expo_hi=0.5, eff=0.8, cost=0.01, coverage=1)

#to compare 2 previous scenarios
#rot_plot_resistance(dfrot, df_resanother=dfseq, plot_refuge=FALSE)

@@

\newpage
###costs of resistance

Costs increased to 5%.

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="A sequence with resistance cost of 0.05."}

dfseq <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=0, expo_hi=0.5, eff=0.8, cost=0.05, coverage=1)

@@

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="A 1 year rotation with resistance cost of 0.05."}

dfrot <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=10, expo_hi=0.5, eff=0.8, cost=0.05, coverage=1)

#to compare 2 previous scenarios
#rot_plot_resistance(dfrot, df_resanother=dfseq, plot_refuge=FALSE)

@@

\newpage
###immigration from untreated areas

Even without costs of resistance we might expect resistance frequencies to decline when an insecticide is not in use due to immigration from nearby areas that were not treated originally. This is shown in the following two plots where a neighbouring untreated area ('refuge') and exchange of migrants is represented. In this example the treated and untreated areas are set to the same size and migration at 1%. The time it takes for both insecticides to reach the resistance threshold of 50% is still very similar for a sequence and a rotation.

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="A sequence with 1 percent dispersal between treated and untreated areas of equal size (no resistance costs)."}

dfr <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=0, expo_hi=0.5, eff=0.8, cost=0, coverage=0.5, migration = 0.01)

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="A 1 year rotation with 1 percent dispersal between treated and untreated areas of equal size (no resistance costs)."}

dfr <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=10, expo_hi=0.5, eff=0.8, cost=0, coverage=0.5, migration = 0.01)

\newpage
###costs and immigration

Including both costs of resistance and immigration from an untreated area, still results in little difference between a sequence and rotation in terms of time-to-resistance to both insecticides.

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="A sequence with dispersal between treated and untreated areas and resistance cost of 0.005."}

#upped max_gen because migration and cost decrease rise in resistance
dfr <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=0, expo_hi=0.5, eff=0.8, cost=0.005, coverage=0.5, migration = 0.01)

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="A 1 year rotation with with dispersal between treated and untreated areas and resistance cost of 0.005."}

dfr <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=10, expo_hi=0.5, eff=0.8, cost=0.005, coverage=0.5, migration = 0.01)

\newpage
###dominance

In all the preceeding plots dominance of resistance and dominance of cost were set to 0.5. Here we check whether changing the dominance of cost between 0 and 1 has an effect on rotations. It does not appear to change time to the 50% resistance threshold, on the following page we try increasing the threshold.

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="A 1 year rotation with a resistance cost of 0.005 and dominance of cost set to 0 (recessive)."}

dfr <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=10, expo_hi=0.5, eff=0.8, cost=0.005, coverage=1, migration = 0, dom_cos = 0)

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="A 1 year rotation with a resistance cost of 0.005 and dominance of cost set to 1 (dominant)."}

dfr <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=10, expo_hi=0.5, eff=0.8, cost=0.005, coverage=1, migration = 0, dom_cos = 1)

\newpage
###dominance of cost effects with increased resistance threshold (90%)

Still little difference between sequence and rotation.

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="dominance of cost 0"}

dfr <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=0, expo_hi=0.5, eff=0.8, cost=0.005, coverage=1, migration = 0, dom_cos = 0, rot_criterion = 0.9, start_freqs = 0.001)

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="dominance of cost 0."}

dfr <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=10, expo_hi=0.5, eff=0.8, cost=0.005, coverage=1, migration = 0, dom_cos = 0, rot_criterion = 0.9, start_freqs = 0.001)

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="dominance of cost 1."}

dfr <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=0, expo_hi=0.5, eff=0.8, cost=0.005, coverage=1, migration = 0, dom_cos = 1, rot_criterion = 0.9, start_freqs = 0.001)

```{r, eval=TRUE, fig.height=2.5, fig.width=7, fig.show='hold', fig.align='left', fig.cap="dominance of cost 1."}

dfr <- run_rot(n_insecticides=2, max_gen = 200, rot_interval=10, expo_hi=0.5, eff=0.8, cost=0.005, coverage=1, migration = 0, dom_cos = 1, rot_criterion = 0.9, start_freqs = 0.001)
